---
title: "Add user defined TDF data to SIDER"
author: "Andrew Jackson"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup}
library(SIDER)
library(MixSIAR)
library(dplyr)

```
In this example we will illustrate how you can add additional data for known TDFs to the dataset and run SIDER. We also compare the results obtained from SIDER should the new species be added with the case when it was omitted entirely. In compiling the dataset for SIDER we applied strict rules on the data, including that he TDFs be from a strictly controlled feeding diet. In many cases this excluded data from studies where the TDFs were inferred indirectly from an experiemental manipulation such as when [Derbridge et al (2015)](https://doi.org/10.1371/journal.pone.0119940) inferred TDFs for wolves (**Canis lupus**) using a modified Stable Isotope Mixing Model in which the TDFs themselves were estimated. Here we will use SIDER to estimate the TDF for wolf hair when wolves are absent from the dataset, and when we include the Derbridge et al (2015) data.

It may seem like over-kill to run SIDER for a species for which you have data, and the obvious question is "why would you do this?". It comes down to how much you trust the TDFs for your species. If you are confident that they most fairly and acccurately reflect the consumers in your mixing model, then you should use them. If however they are from a different population, or at a different time, or they lack the number of replicates etc... then in all likelihood you are not as certain of them as you might be. By adding the data you do have to SIDER and then running SIDER with that same species and tissue missing, then you are getting an estimate that is more informed by the dataset total. Of course it will be weighted more heavily towards the same species data in the dataset owing to the phylogenetic correlation structure in the models, but you will also get estimates of the uncertainty that might be lacking if your data comprise only the mean estimates.



## SIDER data without wolves

```{r SIDER_tdfs}

# Read in the data
SIDER.data <- read.csv(file = system.file("extdata", 
                                          "SIDER_data.csv", 
                                          package = "SIDER"), 
                       header = TRUE,
                       stringsAsFactors = FALSE)

# View the first 10 rows of the data frame
head(SIDER.data)

# Read in the phylogenetic information
# The mammal trees
mammal_trees <- ape::read.tree(system.file("extdata", 
                                      "3firstFritzTrees.tre", 
                                      package = "SIDER"))
# The bird trees
bird_trees   <- ape::read.tree(system.file("extdata", 
                                      "3firstJetzTrees.tre", 
                                      package = "SIDER"))

# Combine them together using the tree.bind function from the mulTree package
combined_trees <- mulTree::tree.bind(x = mammal_trees, 
                                     y = bird_trees, 
                                     sample = 2, 
                                     root.age = 250)

# Checking the data for the species we want to estimate TEF for (Meles meles)
new.data.wolf <- recipeSider(species = "Canis_lupus", 
                             habitat = "terrestrial", 
                             taxonomic.class = "mammalia", 
                             tissue = "hair", 
                             diet.type = "carnivore", 
                             tree = combined_trees)

# prepare the carbon model
tdf_data_c <- prepareSider(new.data.wolf, 
                          isotope_data, 
                          combined_trees, 
                          "carbon")

# prepare the nitrogen model
tdf_data_n <- prepareSider(new.data.wolf, 
                          isotope_data, 
                          combined_trees, 
                          "nitrogen")

# formulae for both
formula.c <- delta13C ~ diet.type + habitat
formula.n <- delta15N ~ diet.type + habitat

# common random structure for both
random.terms <- ( ~ animal + species + tissue)

prior <- list(R = list(V = 1, nu=0.002), 
              G = list(G1=list(V = 1, nu=0.002),
                       G2=list(V = 1, nu=0.002), 
                       G3=list(V = 1, nu=0.002)))

# model run settings
# *Might* want these to be longer runs.
nitt <- c(220000)
burnin <- c(20000)
thin <- c(100)
parameters <- c(nitt, thin, burnin)
no.chains <- c(3)

# convergence settings
convergence =  c(1.1)
ESS = c(1000)


# ************** VIGNETTE *********************
# These dont work in notebooks
# origwd <- getwd() # get the current, or original working directory.
# # setwd(tempdir()) # 
# setwd("/tmp")
# ***********************************


TDF_est_out.c <- imputeSider(mulTree.data = tdf_data_c, 
                           formula = formula.c, 
                           random.terms = random.terms,
                           prior = prior, 
                           output = "wolves_sider_c_run",
                           parameters = parameters,
                           chains = no.chains, 
                           convergence =  convergence, 
                           ESS = ESS)

TDF_est_out.n <- imputeSider(mulTree.data = tdf_data_n, 
                           formula = formula.n, 
                           random.terms = random.terms,
                           prior = prior, 
                           output = "wolves_sider_n_run",
                           parameters = parameters,
                           chains = no.chains, 
                           convergence =  convergence, 
                           ESS = ESS)


# **************** VIGNETTE *******************
# ? set back for the vignette
# setwd(origwd)
# ***********************************

###Now lets have a look at the files imputeSider has saved to the current working directory
# list.files(pattern = "test_c_run")

```


## Sider data with wolves

Much of the required data has already been read in and objects created in the code above, so all that remains is that we add the wolf data from [Derbridge et al (2015)](https://doi.org/10.1371/journal.pone.0119940) and re-run the model.

```{r SIDER_tdfs_wolf_in}


# add the wolf data by binding the new data onto the included SIDER dataset
SIDER.data.wolf <- rbind(SIDER.data, data.frame(species = "Canis_lupus", 
                                                habitat = "terrestrial", 
                                                taxonomic.class = "mammalia",
                                                tissue = "hair", 
                                                diet.type =  "carnivore",
                                                source.iso.13C = NA,
                                                source.iso.15N = NA, 
                                                delta13C = 1.97,
                                                delta15N = 3.04))

# check it worked
tail(SIDER.data.wolf)


# Checking the data for the species we want to estimate TEF for (Canis lupus)
new.data.wolf_2 <- recipeSider(species = "Canis_lupus", 
                             habitat = "terrestrial", 
                             taxonomic.class = "mammalia", 
                             tissue = "hair", 
                             diet.type = "carnivore", 
                             tree = combined_trees)

# prepare the carbon model
tdf_data_c <- prepareSider(new.data.wolf_2, 
                          isotope_data, 
                          combined_trees, 
                          "carbon")

# prepare the nitrogen model
tdf_data_n <- prepareSider(new.data.wolf_2, 
                          isotope_data, 
                          combined_trees, 
                          "nitrogen")


# ************** VIGNETTE *********************
# These dont work in notebooks
# origwd <- getwd() # get the current, or original working directory.
# # setwd(tempdir()) # 
# setwd("/tmp")
# ***********************************


TDF_est_in.c <- imputeSider(mulTree.data = tdf_data_c, 
                           formula = formula.c, 
                           random.terms = random.terms,
                           prior = prior, 
                           output = "wolves_2_sider_c_run",
                           parameters = parameters,
                           chains = no.chains, 
                           convergence =  convergence, 
                           ESS = ESS)

TDF_est_in.n <- imputeSider(mulTree.data = tdf_data_n, 
                           formula = formula.n, 
                           random.terms = random.terms,
                           prior = prior, 
                           output = "wolves_2_sider_n_run",
                           parameters = parameters,
                           chains = no.chains, 
                           convergence =  convergence, 
                           ESS = ESS)


# **************** VIGNETTE *******************
# ? set back for the vignette
# setwd(origwd)
# ***********************************

###Now lets have a look at the files imputeSider has saved to the current working directory
# list.files(pattern = "test_c_run")

```

## Compare the output from both models

In this example, there is very little difference in the two models.

```{r compare-models}

df_wide <- data.frame(DC_in  = as.numeric(TDF_est_in.c$tdf_global),
                      DC_out = as.numeric(TDF_est_out.c$tdf_global), 
                      DN_in  = as.numeric(TDF_est_in.n$tdf_global),
                      DN_out = as.numeric(TDF_est_out.n$tdf_global)
                      )

summary(df_wide)

# tidy to long format for nice plotting
df_long <- tidyr::gather(df_wide, model, TDF)

smrys <- df_long %>% group_by(model) %>% 
  summarise(mean = mean(TDF), sd = sd(TDF))

ggplot(smrys, aes(model, mean)) + 
  geom_errorbar(data = smrys, 
                mapping = aes(ymin = mean - sd, ymax = mean + sd)) + 
  geom_point() + 
  scale_y_continuous(name="TDF")

```







