---
title: "SIDER to MixSIAR pipeline example"
author: "Andrew Jackson"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup}
library(MixSIAR)
library(ggplot2)

```


In this example we illustrate how to include a SIDER analysis into a workflow for a mixing model anlaysis using MixSIAR.

We will take a single pack from the Wolves example in MixSIAR [**_REF_**] and run a simple mixing model using the default TDFs included with the example which were derived using **_XXX_** and compare that with the results obtained using SIDER to estimate them. For this simplified example we use only wolf pack 1 located in region 1 and so omit any factors. The example will work similarly for the full analysis


```{r load-and-prune-data}

# Replace the system.file call with the path to your file
# wolves.filename <- system.file("extdata", "wolves_consumer.csv", 
#                                package = "SIDER")

# wolves.filename <- paste0(getwd(), "/inst/extdata/wolves_consumer.csv")
mix.filename <- "../inst/extdata/wolves_consumer.csv"

# Load the mixture/consumer data
mix <- load_mix_data(filename = mix.filename, 
                     iso_names = c("d13C","d15N"), 
                     factors = NULL, 
                     fac_random = NULL, 
                     fac_nested = NULL, 
                     cont_effects = NULL)


```


Now read in the source data.
```{r load-sources}
# Replace the system.file call with the path to your file
# source.filename <- system.file("extdata", "wolves_sources.csv", package = "SIDER")
source.filename <- "../inst/extdata/wolves_sources.csv"

# Load the source data
source <- load_source_data(filename=source.filename,
                           source_factors=NULL, 
                           conc_dep=FALSE, 
                           data_type="means", 
                           mix)
```



Load the TDF values used in the original data

```{r load-original-TDFs}
# Replace the system.file call with the path to your file
# discr.filename <- system.file("extdata", "wolves_discrimination.csv", package = "MixSIAR")
discr.filename <- "../inst/extdata/wolves_discrimination.csv"

# Load the discrimination/TDF data
discr <- load_discr_data(filename=discr.filename, mix)

```


## Fit the basic MixSIAR model using the original TDFs

```{r orig-tdf-model, results = 'hide', message=FALSE}
# Write the JAGS model file
model_filename <- "MixSIAR_model.txt"   # Name of the JAGS model file
resid_err <- TRUE
process_err <- FALSE
write_JAGS_model(model_filename, resid_err, process_err, mix, source)


jags.1 <- run_model(run="short", mix, source, discr, model_filename, 
                    alpha.prior = 1, resid_err, process_err)

```

## Use "common" errors on the TDFs

```{r tdf_with_error, results = 'hide', message=FALSE}

discr2 <- discr
discr2$sig2[,"SDd13C"] <- 0.63
discr2$sig2[,"SDd15N"] <- 0.74

jags.2 <- run_model(run="short", mix, source, discr2, model_filename, 
                    alpha.prior = 1, resid_err, process_err)

```

## Use SIDER to estimate TDFs (discrimination factors)

The wolf isotope samples are described fully in Darimont, C. T., Paquet, P. C. and Reimchen, T. E. (2009), Landscape heterogeneity and marine subsidy generate extensive intrapopulation niche diversity in a large terrestrial vertebrate. Journal of Animal Ecology, 78: 126–133.  [doi](http://dx.doi.ord/10.1111/j.1365-2656.2008.01473.x),

```{r SIDER_tdfs}

# Read in the data
SIDER.data <- read.csv(file = system.file("extdata", 
                                          "SIDER_data.csv", 
                                          package = "SIDER"), 
                       header = TRUE,
                       stringsAsFactors = FALSE)

# View the first 10 rows of the data frame
head(SIDER.data)

# Read in the phylogenetic information
# The mammal trees
mammal_trees <- ape::read.tree(system.file("extdata", 
                                      "3firstFritzTrees.tre", 
                                      package = "SIDER"))
# The bird trees
bird_trees   <- ape::read.tree(system.file("extdata", 
                                      "3firstJetzTrees.tre", 
                                      package = "SIDER"))

# Combine them together using the tree.bind function from the mulTree package
combined_trees <- mulTree::tree.bind(x = mammal_trees, 
                                     y = bird_trees, 
                                     sample = 2, 
                                     root.age = 250)

# Checking the data for the species we want to estimate TEF for (Meles meles)
new.data.test <- recipeSider(species = "Canis_lupus", 
                             habitat = "terrestrial", 
                             taxonomic.class = "mammalia", 
                             tissue = "hair", 
                             diet.type = "carnivore", 
                             tree = combined_trees)

# prepare the carbon model
tdf_data_c <- prepareSider(new.data.test, 
                          isotope_data, 
                          combined_trees, 
                          "carbon")

# prepare the nitrogen model
tdf_data_n <- prepareSider(new.data.test, 
                          isotope_data, 
                          combined_trees, 
                          "nitrogen")

# formulae for both
formula.c <- delta13C ~ diet.type + habitat
formula.n <- delta15N ~ diet.type + habitat

# common random structure for both
random.terms <- ( ~ animal + species + tissue)

prior <- list(R = list(V = 1, nu=0.002), 
              G = list(G1=list(V = 1, nu=0.002),
                       G2=list(V = 1, nu=0.002), 
                       G3=list(V = 1, nu=0.002)))

# model run settings
# *Might* want these to be longer runs.
nitt <- c(220000)
burnin <- c(20000)
thin <- c(100)
parameters <- c(nitt, thin, burnin)
no.chains <- c(2)

# convergence settings
convergence =  c(1.1)
ESS = c(1000)


# ************** VIGNETTE *********************
# These dont work in notebooks
# origwd <- getwd() # get the current, or original working directory.
# # setwd(tempdir()) # 
# setwd("/tmp")
# ***********************************


TDF_est.c <- imputeSider(mulTree.data = tdf_data_c, 
                         formula = formula.c, 
                         random.terms = random.terms,
                         prior = prior, 
                         output = "wolves_sider_c_run",
                         parameters = parameters,
                         chains = no.chains, 
                         convergence =  convergence, 
                         ESS = ESS)

TDF_est.n <- imputeSider(mulTree.data = tdf_data_n, 
                         formula = formula.n, 
                         random.terms = random.terms,
                         prior = prior, 
                         output = "wolves_sider_n_run",
                         parameters = parameters,
                         chains = no.chains, 
                         convergence =  convergence, 
                         ESS = ESS)


# **************** VIGNETTE *******************
# ? set back for the vignette
# setwd(origwd)
# ***********************************

###Now lets have a look at the files imputeSider has saved to the current working directory
# list.files(pattern = "test_c_run")

```

Take a look at the summaries of the SIDER runs. First $\Delta^{13}$C.

```{r summarise-sider-carbon}
# Calculate summary statistics of the posterior. 
# Specifically, the mean and standard deviation would be
# taken from here and used in a mixing model analysis using 
# MixSIAR, MixSIR or SIAR for example.
summary(TDF_est.c$tdf_global)

# Credible intervals and the mode of the posterior are obtained 
# using the hdrcde package
hdrcde::hdr(TDF_est.c$tdf_global, prob = c(50, 95, 99))

# You can also create density plots of the posterior
coda::densplot(TDF_est.c$tdf_global)

```

Second, $\Delta^{15}$N.

```{r summarise-sider-nitrogen}
# Calculate summary statistics of the posterior. 
# Specifically, the mean and standard deviation would be
# taken from here and used in a mixing model analysis using 
# MixSIAR, MixSIR or SIAR for example.
summary(TDF_est.c$tdf_global)

# Credible intervals and the mode of the posterior are obtained 
# using the hdrcde package
hdrcde::hdr(TDF_est.c$tdf_global, prob = c(50, 95, 99))

# You can also create density plots of the posterior
coda::densplot(TDF_est.c$tdf_global)

```

Now incorporate the new TDF data in a mixsiar model.

```{r sider-to-mixsiar}

# means and standard deviations of the SIDER estimates
mu.c <- mean(TDF_est.c$tdf_global)
mu.n <- mean(TDF_est.n$tdf_global)

sd.c <- sd(TDF_est.c$tdf_global)
sd.n <- sd(TDF_est.n$tdf_global)

tdf_sider <- list()
tdf_sider$mu <- matrix(c(rep(mu.c, 3), rep(mu.n, 3)), ncol = 2, nrow = 3,
                       dimnames = list(source$source_names, mix$MU_names))
tdf_sider$sd <- matrix(c(rep(sd.c, 3), rep(sd.n, 3)), ncol = 2, nrow = 3,
                       dimnames = list(source$source_names, mix$SIG_names))


```

And finally run mixsiar

```{r sider-mixsiar-model, results = 'hide', message=FALSE}

jags.3 <- run_model(run="short", mix, source, tdf_sider, model_filename, 
                    alpha.prior = 1, resid_err, process_err)

```



## Compare model estimates

```{r compare-model-estimates}

cat("Model 1: original TDFs from the mixsiar example\n")
knitr::kable(jags.1$BUGSoutput$summary)

cat("Model 2: original TDFs from the mixsiar example with extra error\n")
knitr::kable(jags.2$BUGSoutput$summary)

cat("Model 3: SIDER derived TDFs\n")
knitr::kable(jags.3$BUGSoutput$summary)

```

Bundle all the posteriors together using dplyr for plotting with ggplot. Here we focus on the proportion of Deer in the diet of this wolf pack, being the largest constituent of the diet. As we can see, using the TDFs from SIDER increase slightly the estimate of Deer in the diet owing to the slightly higher TDF used for $\Delta^{15}$N. The inclusion of additional error widened the esimated proportion for the "Added Variation" model, but not the SIDER model despite having considerably larger standard deviations: this is most likely due to the proportions moving closer to the boundary at $p_{Deer} = 1$, where there is a natural compressive effect on the distribution.

```{r bundle-output}

post_df <- dplyr::bind_rows(list(as.data.frame(jags.1$BUGSoutput$sims.matrix), 
                                 as.data.frame(jags.2$BUGSoutput$sims.matrix),
                                 as.data.frame(jags.3$BUGSoutput$sims.matrix)),
                            .id = "Model")

# change the names to match the sources
names(post_df) <- c("Model", "Deviance", source$source_names)

# convert to numeric and then factor
post_df$Model <- as.numeric(post_df$Model)
post_df$Model <- factor(post_df$Model, labels = c("Original","Added Variation","SIDER"))

# plot the Deer.
p1 <- ggplot(post_df, mapping = aes(x = Model, y = Deer)) + 
  geom_violin(draw_quantiles = c(0.025, 0.25, 0.5, 0.75, 0.975))
print(p1)
```





